---
title: "Introduction to jamma"
author: "James M. Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jamma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_init, echo=FALSE}
ragg_png <- function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(
  collapse=TRUE,
  warning=FALSE,
  message=FALSE,
  comment="#>",
  fig.path="jammaVignette-"
);
library(jamma);
library(jamba);
```

# Introduction to jamma

The jamma package is primarily intended to produce MA-plots,
a special form of Bland-Altman plot applied to genomics or omics data,
typically representing log2-transformed expression measurements form
DNA or RNA microarrays, RNA-seq counts, or other expression measurements.

MA-plots have great utility as a data quality control (QC) methodology,
as they effectively depict several aspects of data directly related to
common downstream statistical assumptions. They can make it easy to recognize
an obvious sample failure, or similarly to identify when one or more samples
are not properly normalized, relative to downstream analysis steps.

That said, MA-plots often suffer from ineffective display of tens of
thousands of datapoints, and so this package was motivated initially simply
by the desire to display the density of datapoints rather than each individual
point. It evolved to provide specific benefits, driven by a large number
of analysis projects across several different platform technologies.

The jamma package also contains helper functions used to create and manipulate
MA-plots which have been quite useful. This vignette demonstrates MA-plots,
the various alternate ways of displaying MA-plot data, and the usefulness
or the helper functions.

## MA-plot content

MA-plots show difference in log2 signal from average, as a function of
average log2 signal. Typically one creates MA-plots to compare multiple
samples, but one can compare sample groups using group mean values.
The typical MA-plot for our purpose will display an individual sample
in each panel.

The visual distribution of points is intended to show the spread of
points from low to high expression along the x-axis, and the deviation
from mean along the y-axis. The relationship of signal (x-axis) to
noise (y-axis) is a visual indicator of several characteristics of the
data, and impacts the choice and effectiveness of numerous downstream
statistical tests. The review of numerous panels of MA-plots can also
be an effective way to judge the consistency of data from the samples
and sample groups.

### Example MA-plot

A reasonable example MA-plot can be created using data from the affydata
package.

```{r Dilution, results="hide", fig.height=8, fig.width=8}
if (suppressPackageStartupMessages(require(affydata))) {
   data(Dilution);
   if (max(exprs(Dilution), na.rm=TRUE) > 100) {
      exprs(Dilution) <- log2(1 + exprs(Dilution));
   }
   jammaplot(exprs(Dilution));
}
```

## Interpretation of MA-plots

MA-plots display several really useful features which tie directly into
many of the assumptions of downstream statistical tools. Often, when something
appears puzzling in downstream analysis results, MA-plots are a great place
to look into what might be causing the issues.

For example, in the MA plots above, it is clear the data has not been
normalized. The evidence is that the points are not horizontal, centered
with y-axis value of zero.
